---
title: "Lab 4"
author: "Jayson De La O"
format:
  html:
    embed-resources: true
---


Steps
1. Read in the data
First download and then read in with data.table::fread()

```{r}
library(tidyverse)
library(data.table)
library(leaflet)
library(dplyr)
library(devtools)
library(R.utils)
```


```{r}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")
```
Remove temperatures less than -17C
Make sure there are no missing data in the key variables coded as 9999, 999, etc
Generate a date variable using the functions as.Date() (hint: You will need the following to create a date paste(year, month, day, sep = "-")).
Using the data.table::week function, keep the observations of the first week of the month.
Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.
Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
Create a categorical variable for elevation as in the lecture slides
```{r}
summary(met$temp)
met <- met[temp>=-17]
summary(met$temp)
summary(met)

met[met$elev==9999.0] <- NA
```

```{r}
met$date<-as.Date(paste(met$year, met$month, met$day, sep = "-"))
```

```{r}
met$week <- data.table::week(met$date)
tail(met)
start <- as.Date("2019-08-01")
end <- as.Date("2019-08-07")
```
```{r}
metfw<- met[met$date >= start & met$date <=end]
metfw
```

```{r}
metfw_avg<- metfw[,.(
   temp     = mean(temp,na.rm=TRUE),
  rh       = mean(rh,na.rm=TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  vis.dist = mean(vis.dist,na.rm=TRUE),
  dew.point  = mean(dew.point,na.rm=TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
),by =c("USAFID")]

metfw_avg

```

```{r}
metfw_avg$region <- ifelse(metfw_avg$lon > -98 & metfw_avg$lat > 39.71,"NE",
                       ifelse(metfw_avg$lon > -98 & metfw_avg$lat <39.71,"NW",
                              ifelse(metfw_avg$lon < -98 & metfw_avg$lat > 39.71,"SE",
                       ifelse(metfw_avg$lon < -98 & metfw_avg$lat <39.71,"SW",NA))))
                              
unique(metfw_avg$region)
```
The graphs shows that the south (SW/SE) have longer range of dew points, while the North (NW/NE) has a smaller range, but higher overall dew points.
Wind speeds seems to be relative equal throughout the regions, but seems to be slightly higher in the south.
```{r}
metfw_avg[,elev_cat := fifelse(elev > 252, "high","low")]


```
```{r}
metfw_avg[!is.na(wind.sp) & !is.na(dew.point) & !is.na(region)] %>%
  ggplot()+
  geom_violin(mapping=aes(x=1,y=dew.point,fill=region))+
  facet_wrap(~region,nrow=1)

metfw_avg[!is.na(wind.sp)& !is.na(dew.point) & !is.na(region)] %>%
  ggplot()+
  geom_violin(mapping=aes(x=1,y=wind.sp,fill=region))+
  facet_wrap(~region,nrow=1)


```
Wind speed and dew point seem to be relatively linear for each of the regions. Showing that wind speed has an affect on dew point.

```{r}
metfw_avg[!is.na(wind.sp)& !is.na(dew.point) & !is.na(region)] %>%
 ggplot( aes(x=dew.point,y=wind.sp,color=region ))+
  geom_jitter(width=.2,height=0)+
  stat_smooth(method=lm,aes(group=region))+
    scale_color_manual(values=c("NE"="blue","SE"="green","NW"="black","SW"="red"))


```

The bar graph shows me that the Northwest has the highest amount of weather stations at low elevations and the souther east has the lowest amount of elevations at low elevations.At high elevations, NE has the highest amount of weather stations and the south east has the lowest amount.
```{r}
metfw_avg[!is.na(elev_cat) & !is.na(region)] %>%
  ggplot(mapping=aes(x=elev_cat,fill=region))+
  geom_bar(position="dodge")+
  scale_fill_brewer(palette="Paired")+
  labs(title="Weather stations -Elev Category by region",x="Elevation Category",y="Number of Weather Stations")
```

Dew point is:  NE=16.81,NW=21.36,SE=11.98, and SW=13.32.
Wind speed is: NE=1.96,NW=1.83,SE=2.90,SW=3.15,

```{r}
metfw_avgmean<- metfw_avg[,.(
   temp     = mean(temp,na.rm=TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  dew.point  = mean(dew.point,na.rm=TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
),by =c("region")]

metfw_avgmean
metfw_avg[!is.na(dew.point) & !is.na(region) & !is.na(wind.sp)] %>%
 ggplot(mapping=aes(y=dew.point,x=region))+
  stat_summary(fun.data = "mean_sdl")+
  stat_summary(geom="errorbar")

metfw_avg[!is.na(dew.point) & !is.na(region) & !is.na(wind.sp)] %>%
 ggplot(mapping=aes(y=wind.sp,x=region))+
  stat_summary(fun.data = "mean_sdl")+
  stat_summary(geom="errorbar")

```

Relative humidity increases near the coasts, but overall it is higher in the East than it is in the West.
```{r}
  

met_avg2 <- met[,.(rh = mean(rh,na.rm=TRUE), lat = mean(lat), lon = mean(lon)),  by=c("USAFID")]
met_avg2 <- met_avg2[!is.na(rh)]

top10<-met_avg2[rank(-met_avg2$rh)<=10]
top10
# Generating a color palette
rh.pal <- colorNumeric(c('darkgreen','goldenrod','brown'), domain=met_avg2$rh)
rh.pal

rhmap <- leaflet(met_avg2) %>% 
  # The looks of the Map
  addProviderTiles('CartoDB.Positron') %>% 
  # Some circles
  addCircles(
    lat = ~lat, lng=~lon,
                                                  # HERE IS OUR PAL!
    label = ~paste0(round(rh,2), ' %'), color = ~ rh.pal(rh),
    opacity = 1, fillOpacity = 1, radius = 500
    ) %>%
  # And a pretty legend
  addLegend('bottomleft', pal=rh.pal, values=met_avg2$rh,
          title='Relative Humidity, %', opacity=1) %>%
  addMarkers(lng=top10$lon,lat=top10$lat)

rhmap
```

```{r}

library(ggforce)

ggplot(metfw_avg, aes(x=elev, y=temp, color = region)) +
  geom_point() +
  facet_zoom(x = Location=="SW")
```

